Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-08 paper/paper.tex
more fixes

The pushdown AAM with an unbounded stack is also described by \citeauthor{van2012systematic}
\cite{van2012systematic}.
We show it here is to maintain consistency of abstract semantics during
transformations -- since the eventual abstract definitional interpreters simply inherit
a precise control-flow from their defining language, we also would like to start from a
pushdown AAM that is also able to precisely match calls and returns.
For readers who are not familiar with pushdown analysis, we provide a detailed discussion
in Section~\ref{pdcfarevisit}.

@@

\textit{Note on Termination.}
Unfortunately, even though other components in the states are finite, this version of AAM with an unbounded
stack may still diverge when analyzing some programs.
This is because the unbounded stack can grow to an arbitrary depth which implies that the state
space is possibly infinite; the analysis may therefore not terminate for all programs if we
simply enumerate reachable states. To see this, consider a program that has two mutually
recursive functions:

@@

At this point, we have obtained a linearized abstract abstract machine.
Intuitively, if we imagine that the classical AAM explores a graph of reachable states,
then the linearized version of AAM flattens the graph in depth-first order to a linear sequence of states.
It has been observed that ``\textit{failure can be replaced by a list of successes}''
\cite{10.1007/3-540-15975-4_33} and that ``\textit{one programmer’s lazy-list
constructor is another programmer’s success and failure continuations}'' \cite{danvy2006a};
the @step@ function in AAM is the list constructor used in a lazy way, and here we turn it
into an explicit data types representing continuation.
The inverse transformation of this step is to apply \citeauthor{10.1007/3-540-15975-4_33}'s
technique that converts the data types representing meta-continuations (failure continuations)
back to lazy list representation, where the elements in the list are plain machine states.

@@

With fusing complete, our AAM appears similar to a ``big-step'' interpreter, although it
still has the machine state representation inside. The inverse transformation of this step
would be splitting the work-list algorithm apart to an individual function.

@@

Refunctionalization transforms first-order data types representing evaluation contexts
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain an abstract interpreter written in extended
continuation-passing style~\cite{Danvy:1990:AC:91556.91622}.
This transformation can be also regarded as making the control flows of the defined
language be part of the control flows of the meta-language.
In this section, we first present a direct refunctionalization of the continuations.
Next, we adopt a caching algorithm to obtain a computable analysis~\cite{darais2017abstracting},
and further prepare the interpreter for transformation to direct-style at a later point.

@@

We have now completely refunctionalized an abstract abstract machine; the two layers
of continuations are represented by higher-order functions. The inverse transformation
of this step is known as \textit{de}functionalization, which converts the higher-order functions
representing continuations back to first-order data
types~\cite{Danvy:2001:DW:773184.773202, Danvy:2008:DIP:1411204.1411206}.

@@

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property in the analysis that precisely models the run-time
call structure of the analyzed program. A pushdown control-flow analysis provides an
as-exact-as-runtime return-flow when analyzing a program, but traditional
control-flow analysis collapses the state space into a finite space and
thus causes imprecise stack modeling.

@@

\textit{Call/Return Matching through Refunctionalization and Caching.}
The refunctionalized AAM with caching has the perfect match on return-flows even though
we do not have a stack model. The higher-order functions representing continuations
already connect all the executions in order.

@@

\begin{itemize}
  \item By replacing the continuations with monads \cite{filinski1994representing},
    we obtain the abstract definitional interpreters in monadic style which are similar to what
    \citet{darais2017abstracting} describe.
  \item By representing the continuation-passing style with delimited control operators
    such as @shift@ and @reset@, we derive a new form of abstract interpreters in direct-style.
    This transformation is the left-inverse of the CPS transform \cite{DBLP:journals/scp/Danvy94, DBLP:conf/lfp/DanvyL92}.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same definitional
    interpreter with pushdown control-flows. \Citet{friedman-ai}, in a tutorial on abstract
    interpreters, use this style to model caches and updates.
    In this case, nondeterminism can be easily handled via @for@ comprehensions.
\end{itemize}

@@

\textit{Defunctionalization and Refunctionalization.}
\citet{Reynolds:72} proposed defunctionalization as a program transformation
technique that can be used to transform higher-order functions to first-order data types.
While the technique is very general, it was first presented as a key step in
transforming higher-order definitional interpreters into their first-order
counterparts.
As a form of generalized closure conversion,
defunctionalization is widely used in compilation and analysis \cite{pottier2006polymorphic,
Eisenberg:2014:PFT:2633357.2633361, fourtounis2014modular, 10.1007/3-540-46425-5_4, consel1993tour}.
Refunctionalization~\cite{Danvy:2006:RW:2171265.2171268} as a
left inverse of defunctionalization, was introduced
by \citeauthor{Danvy:2006:RW:2171265.2171268,DANVY2009534}.

\Citet{Ager:2003:FCE:888251.888254} observed that defunctionalization and refunctionalization
can be used to show the functional correspondence between interpreters and abstract machines
\cite{Ager:2003:FCE:888251.888254}. A number of independently designed small-step semantic artifacts can
be obtained by closure-converting, CPS-transforming, and then defunctionalizing their big-step counterparts,
and vice versa by refunctionalizing \cite{DANVY2009534}.
Such semantic artifacts include the SECD machine \cite{10.1007/11431664_4, Ager:2003:FCE:888251.888254, DanvySECDJ},
the CEK machine \cite{Ager:2003:FCE:888251.888254}, the CLS machine  \cite{Ager:2003:FCE:888251.888254},
Categorical Abstract Machines \cite{Ager:2003:FCE:888251.888254}, lazy abstract machines \cite{AGER2004223},
monadic evaluators \cite{ager2005functional}, as well as languages with richer constructs
\cite{Danvy:2008:DIP:1411204.1411206, danvy2009towards, biernacka2009towards}, etc.

@@

\textit{Lightweight Fusion.}
Lightweight fusion as one of the transformation in our paper was presented by
\citeauthor{Ohori:2007:LFF:1190216.1190241} \cite{Ohori:2007:LFF:1190216.1190241}.
As shown by \citeauthor{DANVY2008100}, lightweight fusion can be used to show the
equivalence between small-step and big-step abstract machines \cite{DANVY2008100}.\\

\textit{Abstract Interpretation and Control-Flow Analysis.}
\citeauthor{cousot1977abstract} discovered abstract interpretation as a sound approach to
approximate a program's run-time behavior \cite{cousot1977abstract}. Control-flow analysis
is one instance of abstract interpretation on higher-order functional programs that can be traced to
Jones \cite{jones1981flow}. \Citet{Shivers:1988:CFA:53990.54007,Shivers:1991:SSC:115865.115884}
introduced $k$-CFA which uses $k$ recent calling contexts as program contours that
differentiate values from different contexts.
A recent formulation of control-flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point
of this paper. Modern programming languages such as Java \cite{might2010resolving} and
Racket \cite{Tobin-Hochstadt:2012:HSE:2384616.2384655} can also be modeled by the abstracting
abstract machines approach. \\

>>>>

The pushdown AAM with an unbounded stack is also described by \citet{van2012systematic}.
We show it here is to maintain consistency of abstract semantics during
transformations -- since the eventual abstract definitional interpreters simply inherit
a precise control-flow from their defining language, we also would like to start from a
pushdown AAM that is also able to precisely match calls and returns.
For readers who are not familiar with pushdown analysis, we provide a detailed discussion
in Section~\ref{pdcfarevisit}.

@@

\textit{Note on Termination.}
Unfortunately, even though other components in the state are finite, this version of AAM with an unbounded
stack may still diverge when analyzing some programs.
This is because the unbounded stack can grow to an arbitrary depth which implies that the state
space is possibly infinite; the analysis may therefore not terminate for all programs if we
simply enumerate reachable states. To see this, consider a program that has two mutually
recursive functions:

@@

At this point, we have obtained a linearized abstract abstract machine.
Intuitively, if we imagine that the classical AAM explores a graph of reachable states,
then the linearized version of AAM flattens the graph in depth-first order to a linear sequence of states.
It has been observed that ``\textit{failure can be replaced by a list of successes}''
\cite{10.1007/3-540-15975-4_33} and that ``\textit{one programmer’s lazy-list
constructor is another programmer’s success and failure continuations}'' \cite{danvy2006a};
the @step@ function in AAM is the list constructor used in a lazy way, and here we turn it
into an explicit data types representing continuation.
The inverse transformation of this step is to apply \citeauthor{10.1007/3-540-15975-4_33}'s
technique that converts the data types representing meta-continuations (failure continuations)
back to a lazy list representation, where the elements in the list are plain machine states.

@@

With fusion complete, our AAM appears similar to a ``big-step'' interpreter, although it
still has the machine state representation inside. The inverse transformation of this step
would be splitting the work-list algorithm apart to an individual function.

@@

Refunctionalization transforms first-order data types representing evaluation contexts
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain an abstract interpreter written in extended
continuation-passing style~\cite{Danvy:1990:AC:91556.91622}.
This transformation can be also regarded as expressing the control flow of the defined
language through control flow of the defining language.
In this section, we first present a direct refunctionalization of the continuations.
Next, we adopt a caching algorithm to obtain a computable analysis~\cite{darais2017abstracting},
and further prepare the interpreter for transformation to direct-style at a later point.

@@

We have now completely refunctionalized an abstract abstract machine; the two layers
of continuations are represented by higher-order functions. The inverse transformation
of this step is known as \textit{de}functionalization, which converts the higher-order functions
representing continuations to first-order data
types~\cite{Danvy:2001:DW:773184.773202, Danvy:2008:DIP:1411204.1411206}.

@@

\textit{The Problem with Return-flows.}
Pushdown control-flow is a property of the analysis that precisely models the run-time
call structure of the analyzed program. A pushdown control-flow analysis provides an
as-exact-as-runtime return-flow when analyzing a program, but traditional
control-flow analysis collapses the state space into a finite space and
thus causes imprecise stack modeling.

@@

\textit{Call/Return Matching through Refunctionalization and Caching.}
The refunctionalized AAM with caching precisely matches return-flows even though
we do not have a stack model. The higher-order functions representing continuations
already connect all the executions in order.

@@

\begin{itemize}
  \item By replacing the continuations with monads \cite{filinski1994representing},
    we obtain the abstract definitional interpreters in monadic style which are similar to what
    \citet{darais2017abstracting} describe.
  \item By representing the continuation-passing style with delimited control operators
    such as @shift@ and @reset@, we derive a new form of abstract interpreters in direct-style.
    This transformation is the left-inverse of the CPS transform \cite{DBLP:journals/scp/Danvy94, DBLP:conf/lfp/DanvyL92}.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same definitional
    interpreter with pushdown control-flow. \Citet{friedman-ai}, in a tutorial on abstract
    interpreters, use this style to model caches and updates.
    In this case, nondeterminism can be easily handled via @for@ comprehensions.
\end{itemize}

@@

\textit{Defunctionalization and Refunctionalization.}
\citet{Reynolds:72} proposed defunctionalization as a program transformation
technique that can be used to transform higher-order functions to first-order data types.
While the technique is very general, it was first presented as a key step in
transforming higher-order definitional interpreters into their first-order
counterparts.
As a form of generalized closure conversion,
defunctionalization is widely used in compilation and analysis \cite{pottier2006polymorphic,
Eisenberg:2014:PFT:2633357.2633361, fourtounis2014modular, 10.1007/3-540-46425-5_4, consel1993tour}.
Refunctionalization as a
left inverse of defunctionalization, was introduced
by \citet{Danvy:2006:RW:2171265.2171268,DANVY2009534}.

\Citet{Ager:2003:FCE:888251.888254} observed that defunctionalization and refunctionalization
can be used to show the functional correspondence between interpreters and abstract machines. A number of independently designed small-step semantic artifacts can
be obtained by closure-converting, CPS-transforming, and then defunctionalizing their big-step counterparts,
and vice versa by refunctionalizing \cite{DANVY2009534}.
Such semantic artifacts include the SECD machine \cite{10.1007/11431664_4, Ager:2003:FCE:888251.888254, DanvySECDJ},
the CEK machine \cite{Ager:2003:FCE:888251.888254}, the CLS machine  \cite{Ager:2003:FCE:888251.888254},
Categorical Abstract Machines \cite{Ager:2003:FCE:888251.888254}, lazy abstract machines \cite{AGER2004223},
monadic evaluators \cite{ager2005functional}, as well as languages with richer constructs
\cite{Danvy:2008:DIP:1411204.1411206, danvy2009towards, biernacka2009towards}, etc.

@@

\textit{Lightweight Fusion.}
Lightweight fusion as one of the transformation in our paper was presented by
\citet{Ohori:2007:LFF:1190216.1190241}.
As shown by \citet{DANVY2008100}, lightweight fusion can be used to show the
equivalence between small-step and big-step abstract machines.\\

\textit{Abstract Interpretation and Control-Flow Analysis.}
\Citet{cousot1977abstract} discovered abstract interpretation as a sound approach to
approximate a program's run-time behavior. Control-flow analysis
is one instance of abstract interpretation on higher-order functional programs that can be traced to \citet{jones1981flow}. \Citet{Shivers:1988:CFA:53990.54007,Shivers:1991:SSC:115865.115884}
introduced $k$-CFA which uses $k$ recent calling contexts as program contours that
differentiate values from different contexts.
A recent formulation of control-flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point
of this paper. The AAM approach has been successfully applied to programming languages such as Java \cite{might2010resolving} and
Racket \cite{Tobin-Hochstadt:2012:HSE:2384616.2384655}. \\
